Approximation as Differentiation: Notes

### 2025-05-20 main/formal-structures.tex
Fill in some fwd refs.

The fundamental approach in CHAD is to consider Grothendieck constructions on indexed categories $T : \cat{C}^\op \to \Cat$. An object of the Grothendieck construction $\int T$ is a pair $(X \in \cat{C}, \partial X \in T(X))$, which we read as pairing a space $X$ of ``points'' from $\cat{C}$ with an associated bundle of tangent spaces. The maps from the ``base'' category $\cat{C}$ are used to interpret the programs we are modelling. The maps in the indexed category are the maps of tangents or approximations, being derivatives and Galois connections respectively.

@@

In these special cases when $\cat{C} = \Set$, the Grothendieck construction is the {\em families} constructions $\Fam(\FinVect)$ and $\Fam(\LatGal)$. As is well known~\cite{lawvere63}, $\Fam(\cat{X})$ for any category $\cat{X}$ is the free coproduct completion of $\cat{X}$. In the case that $\cat{X}$ has a (symmetric) monoidal product, then we always get a symmetric monoidal product on $\Fam(\cat{X})$ (using the Cartesian products from $\Set$). When $\cat{X}$ has all (small) products, and the monoidal product is actually a coproduct, then $\Fam(\cat{X})$ is symmetric monoidal closed. If the coproducts are actually {\em biproducts}, then $\Fam(\cat{X})$ is Cartesian closed, and we can interpret higher-order programs.

\todo{Preceeding paragraph needs references to later in the document where these constructions are detailed in full.}

@@

As sketched in \secref{introduction:galois-slicing}, Galois slicing requires that the approximants of a point
$x$ form a bounded lattice $(X, \meet, \join, \top, \bot)$, with least element $\bot$ representing the
approximation that discards all information about $x$, greatest element $\top$ representing the approximation
that retains all information about $x$, and $\meet$ and $\join$ providing two canonical ways to combine
approximations. Thus rather than working directly in $\PosGal$, we consider the following subcategory instead.

@@

\noindent Right adjoints preserves limits and left adjoints preserves colimits, so for any $f \adj g: X \to Y$
in $\LatGal$, $g$ is a \emph{meet-semilattice homomorphism}, i.e.~preserves the meet-semilattice structure
$(X, \meet, \top)$. Similarly, $f$ is a join-semilattice homomorphism with respect to $(X, \join, \bot)$.

One observation about the (bounded) meet-semilattice homomorphisms from $X$ to $Y$ is that they admit a
meet-semilattice structure themselves: the pointwise lifting of $\meet$ to meet-preserving maps $f, g: X \to
Y$, where $(f \meet g)(x) = f(x) \meet g(x): X \to Y$ also preserves meets, giving a notion of $\meet$ for
morphisms, as does the constant map $\top_{X,Y}: X \to Y$ sending any $x \in X$ to $\top_Y$, and which acts a
unit for $\meet$. Dually the (bounded) join-preserving maps from $Y$ to $X$ have a join-semilattice structure
given pointwise by $\join$ and the constant homomorphism $\bot_{Y,X}: Y \to X$ sending any $y \in Y$ to
$\bot_X$. Since these two constructions are adjoint, $\LatGal$ is enriched in the category $\SemiLat$ of
semilattices and semilattice homomorphisms. The hom-set $\LatGal(X,Y)$ forms a bounded meet semilattice with
unit $\top_{X,Y}$ given by the Galois connection $\bot_{Y,X} \adj \top_{X,Y}: X \to Y$ and meet of Galois
connections $(f \adj g) \meet (f' \adj g') = (f \join f') \adj (g \meet g'): X \to Y$.

@@

\subsubsection{Grothendieck construction for indexed categories}

This pattern of objects and morphisms is exactly the \emph{Grothendieck construction} for a particular indexed
category $T: \Set^\op \to \Cat$, namely the functor which assigns to every set $X$ the functor category $T(X)
= \Func{X}{\LatGal}$. Here the (discrete) category $X$ is a plain set and so $\Func{X}{\LatGal}$ is exactly
the category of $X$-indexed families of bounded lattices, with $X$-indexed families of Galois connections
between them as morphisms. The Grothendieck construction $\Grothendieck{}T$
(\defref{grothendieck:Grothendieck} below) is the category of all $\Set$-indexed families of bounded lattices,
together with morphisms that account for both structure within each family and how families transform under
reindexing.

@@

\begin{definition}[Category of $X$-indexed families]
\label{def:fam:Fam}
For any set $X$ (interpreted as a discrete category) and any category $\cat{C}$, recall that the functor
category $\Func{X}{\cat{C}}$ has as objects all functors $F: X \to \cat{C}$, namely $X$-indexed families of
objects of $\cat{C}$, and as morphisms from $F \to G$, all families of morphisms $\eta(x): F(x) \to G(x)$ in
$\cat{C}$ for every $x \in X$, with naturality trivial because $X$ only has identity morphisms.
\end{definition}

@@

\begin{definition}
\label{def:grothendieck:Grothendieck}
Suppose an indexed category $T: \cat{C}^\op \to \Cat$. The \emph{Grothendieck construction} for $T$ is the
category $\Grothendieck{X}T(X)$ (also written $\Grothendieck{}T$) which has as objects, all pairs $(X,
\partial X)$ of an object $X$ of $\cat{C}$ and an object $\partial X$ of $T(X)$, and as morphisms $(X,
\partial X) \to (Y, \partial Y)$, all pairs $(f, \partial f)$ of morphisms $f: X \to Y$ in $\cat{C}$ and
morphisms $\partial f: \partial X \to T(f)(\partial Y)$ in $T(X)$.
\end{definition}

@@

@@

\begin{definition}[Category of families]
For any category $\cat{C}$ define $\Fam(\cat{C})$ to be the Grothendieck construction
$\Grothendieck{X}\Func{X}{\cat{C}}$.
\end{definition}

@@

\subsection{Biproducts and Semi-Additive Categories}

>>>>

The fundamental approach in CHAD is to consider Grothendieck constructions on indexed categories $T : \cat{C}^\op \to \Cat$. An object of the Grothendieck construction $\int T$~, described in more detail in \secref{Grothendieck}, is a pair $(X \in \cat{C}, \partial X \in T(X))$, which we read as pairing a space $X$ of ``points'' from $\cat{C}$ with an associated bundle of tangent spaces. The maps from the ``base'' category $\cat{C}$ are used to interpret the programs we are modelling. The maps in the indexed category are the maps of tangents or approximations, being derivatives and Galois connections respectively.

@@

In these special cases when $\cat{C} = \Set$, the Grothendieck construction is the {\em families} constructions $\Fam(\FinVect)$ and $\Fam(\LatGal)$~(\secref{Fam}). As is well known~\cite{lawvere63}, $\Fam(\cat{X})$ for any category $\cat{X}$ is the free coproduct completion of $\cat{X}$. In the case that $\cat{X}$ has a (symmetric) monoidal product, then we always get a symmetric monoidal product on $\Fam(\cat{X})$ (using the Cartesian products from $\Set$). When $\cat{X}$ has all (small) products, and the monoidal product is actually a coproduct, then $\Fam(\cat{X})$ is symmetric monoidal closed. If the coproducts are actually {\em biproducts}~(\secref{biproducts}), then $\Fam(\cat{X})$ is Cartesian closed, and we can interpret higher-order programs.

@@

As sketched in \secref{introduction:galois-slicing}, for Galois slicing we want a setting where the
approximants of a point $x$ form a bounded lattice $(X, \meet, \join, \top, \bot)$, with least element $\bot$
representing the approximation that discards all information about $x$, greatest element $\top$ representing
the approximation that retains all information about $x$, and $\meet$ and $\join$ providing two canonical ways
to combine approximations. Thus rather than working directly in $\PosGal$, we consider the following
subcategory instead.

@@

\noindent Right adjoints preserves limits and left adjoints preserves colimits, so for any $f \adj g: X \to Y$
in $\LatGal$, $g$ is a (bounded) \emph{meet-semilattice homomorphism}, i.e.~preserves the meet-semilattice
structure $(X, \meet, \top)$. Similarly, $f$ is a join-semilattice homomorphism with respect to $(X, \join,
\bot)$.

The meet-semilattice homomorphisms from $X$ to $Y$ form a meet-semilattice themselves, with $\meet$ on
homomorphisms defined pointwise, so that $(f \meet g)(x) = f(x) \meet g(x): X \to Y$, and the constant map
$\top_{X,Y}: X \to Y$ sending any $x \in X$ to $\top_Y$ providing the unit (and both preserving meets). Dually
the join-preserving maps from $Y$ to $X$ have a join-semilattice structure given pointwise by $\join$ and the
constant homomorphism $\bot_{Y,X}: Y \to X$ sending any $y \in Y$ to $\bot_X$. Since these two constructions
come in adjoint pairs, $\LatGal$ is enriched in the category $\SemiLat$ of semilattices and semilattice
homomorphisms. The hom-set $\LatGal(X,Y)$ forms a bounded meet semilattice with unit $\top_{X,Y}$ given by the
Galois connection $\bot_{Y,X} \adj \top_{X,Y}: X \to Y$ and meet of Galois connections $(f \adj g) \meet (f'
\adj g') = (f \join f') \adj (g \meet g'): X \to Y$.

@@

\subsubsection{Grothendieck construction for indexed categories}
\label{sec:Grothendieck}

This pattern of objects and morphisms is exactly the \emph{Grothendieck construction} for a particular indexed
category $T: \Set^\op \to \Cat$, namely the functor which assigns to every set $X$ the functor category $T(X)
= \Func{X}{\LatGal}$. Here the (discrete) category $X$ is a plain set and so $\Func{X}{\LatGal}$ is exactly
the category of $X$-indexed families of bounded lattices, with $X$-indexed families of Galois connections
between them as morphisms. The Grothendieck construction $\Grothendieck{}T$~(\defref{Grothendieck} below) is
the category of all $\Set$-indexed families of bounded lattices, together with morphisms that account for both
structure within each family and how families transform under reindexing.

@@

\begin{definition}[Category of $X$-indexed families]
For any set $X$ (interpreted as a discrete category) and any category $\cat{C}$, recall that the functor
category $\Func{X}{\cat{C}}$ has as objects all functors $F: X \to \cat{C}$, namely $X$-indexed families of
objects of $\cat{C}$, and as morphisms from $F \to G$, all families of morphisms $\eta(x): F(x) \to G(x)$ in
$\cat{C}$ for every $x \in X$, with naturality trivial because $X$ only has identity morphisms.
\end{definition}

@@

\begin{definition}
\label{def:Grothendieck}
Suppose an indexed category $T: \cat{C}^\op \to \Cat$. The \emph{Grothendieck construction} for $T$ is the
category $\Grothendieck{X}T(X)$ (also written $\Grothendieck{}T$) which has as objects, all pairs $(X,
\partial X)$ of an object $X$ of $\cat{C}$ and an object $\partial X$ of $T(X)$, and as morphisms $(X,
\partial X) \to (Y, \partial Y)$, all pairs $(f, \partial f)$ of morphisms $f: X \to Y$ in $\cat{C}$ and
morphisms $\partial f: \partial X \to T(f)(\partial Y)$ in $T(X)$.
\end{definition}

@@

\subsubsection{Category of families}
\label{sec:Fam}

@@

\begin{definition}[Category of families]
\label{def:Fam}
For any category $\cat{C}$ define $\Fam(\cat{C})$ to be the Grothendieck construction
$\Grothendieck{X}\Func{X}{\cat{C}}$.
\end{definition}

@@

\subsection{Biproducts and Semi-Additive Categories}
\label{sec:biproducts}
